---
title: "Stat 435 HW1"
author: "Abigail Kubli (11885701)"
header-includes:
- \usepackage{bbm}
- \usepackage{amssymb}
- \usepackage{amsmath}
- \usepackage{graphicx}
- \usepackage{natbib}
- \usepackage{float}
- \floatplacement{figure}{H}
output:
  pdf_document: default
  html_document:
    df_print: paged
fontsize: 11pt
---

```{r, echo=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

# General guidelines

This HW covers contents related to simple linear regression and multiple linear regression.

Please show your work in order to get points. Providing correct answers without supporting details does not receive full credits. The homework assignment also trains your programming skills. So, please complete both the conceptual exercises and applied exercises. It is absolutely not OK to just submit your codes only. This will result in a considerable loss of points on your assignments or projects.

For an assignment or project, you DO NOT have to submit your answers or reports using typesetting software. However, your answers must be well organized and well legible for grading. Specifically, if you are not able to knit a .Rmd/.rmd file into an output file such as a .pdf, .doc, .docx or .html file that contains your codes, outputs from your codes, your interpretations on the outputs, and your answers in text (possibly with math expressions), you can organize your codes, their outputs and your answers in a document. Regardless of if you are using typesetting software or not, please organize your work in the format given below:

```         
Problem or task or question ... 
Codes ...
Outputs ...
Your interpretations ...
```

Please upload your answers in a document to the course space where assignment or project is provided.

# Conceptual exercises

1.  Consider a simple linear regression.

  (a) Provide the formulae for the *least squares estimate (LSE)* of the coefficients, based on observations $y_i,i=1,\ldots,n$ for the response $Y$ and observations $x_i,i=1,\ldots,n$ for the predictor $X$.

  Least Squares estimates:

  Slope: $\hat{\beta_1} = \frac{\sum_{i=1}^{n}(x_i-\bar{x})(y_i-\bar{y})} {\sum_{i=1}^{n}(x_i-\bar{x})^2}$ 

  Intercept: $\hat{\beta_0} = \bar{y} - \hat{\beta_1} \bar{x}$ 

  This creates an estimated regression line of $\hat{Y} = \hat{\beta_0} + \hat{\beta_1} X$ modeled off of $Y = \beta_0 + \beta_1 X + \varepsilon$

  (b) Are these estimated coefficients unbiased?

  Yes with the assumption that the standard linear regression assumptions are satisfied, $E(\varepsilon_i|X_i) = 0$ 

  Then $E(\hat{\beta_0}) = \beta_0$ and $E(\hat{\beta_1}) = \beta_1$ 

  So both the estimated slope and intercept are unbiased estimators of their true coefficients. On average the sample estimates will not systematically over or underestimate the true population coefficients.

  (c) Provide the formulae for the standard errors of these estimated coefficients, and state the conditions under which the formulae are valid.

  $SE(\hat{\beta}_1) = \frac{\sigma} {\sqrt{\sum_{i=1}^{n}(x_i-\bar{x})^2}}$ 

  $\hat{\sigma} = \sqrt{\frac{RSS} {n-2}}$

  Thus $SE(\hat{\beta}_1) = \frac{\hat{\sigma}} {\sqrt{\sum_{i=1}^{n}(x_i-\bar{x})^2}}$

  $SE(\hat{\beta}_0) = \hat{\sigma} \sqrt{\frac{1} {n} + \frac{\bar{x}^{2}} {\sum_{i=1}^{n}(x_i-\bar{x})^2}}$

  Conditions:

  The relationship between X & Y is linear

  The error terms have a mean zero

  The error terms are independent

  And the error terms have constant variance: $Var(\varepsilon_i) = \sigma^{2}$

  (d) Explain how the sample variance of the observations $x_i,i=1,\ldots,n$ for $X$ affects the standard errors of the estimated intercept and slope.
  
  We can use the estimated standard deviation of the regression line $\hat{\sigma} = \sqrt{\frac{RSS} {n-2}}$ altered to $\hat{\sigma}^{2} = \frac{RSS} {n-2}$
  
  With the affects of $X$ we can say that the sample variance of $X = s^{2}_X = \frac{1} {n-1} \sum_{i=1}^{n}(x_i - \bar{x}^{2})$. This measures how spread out the $X$ observations are.
  
  That equation lets us write $SE(\hat{\beta}_1) = \frac{\hat{\sigma}} {\sqrt{(n-1)s^{2}_X}}$.
  
  Thus a larger $s^{2}_X$ leads to a smaller $SE(\hat{\beta_1})$ and a smaller $s^{2}_X$ leads to a larger $SE(\hat{\beta_1})$. When the observations of $X$ are more spread out, it's easier to estimate the slope precisely. This also affects the intercept as $SE(\hat{\beta_0}) = \hat{\sigma} \sqrt{\frac{1} {n} + \frac{\bar{x}^{2}} {(n-1)s^{2}_X}}$, so a greater variation in $X$ generally reduces the second term which then reduces the uncertainty in $\hat{\beta_0}$.
  
  (e) Give a test on if the slope in the model is zero, by stating the null hypothesis, the statistic with its degrees of freedom, and the decision rule. Under what conditions will this test work well?

  $H_0: \beta_1 = 0$ and $H_A: \beta_1 \neq 0$
  
  At significance level $\alpha$, we reject $H_0$ when $|t|> t_{\alpha/2, n-2}$, where n-2 is the degrees of freedom and $t = \frac{\hat{\beta_1}}{SE(\hat{\beta_1})}$. Otherwise, we fail to reject $H_0$.
  
  This works well when the linear regression assumptions are reasonably satisfied (these assumptions/conditions were explained in part (c)). Having approximately normal errors are particularly important for a $t$-test.

  (f) In the simple linear regression model $Y=\beta_0 + \beta_1 X + \varepsilon$, $\varepsilon$ is the random error term with unknown standard deviation $\sigma$. Provide an estimator for $\sigma$, and describe how the sample size $n$ of the observations affects the accuracy of this estimator.
  
  The error standard deviation $\sigma$ is estimated using $RSS = \sum_{i=1}^{n}(y_i-\bar{y})^{2}$ creating $\hat{\sigma} = \sqrt{\frac{RSS}{n-2}} = \sqrt{\frac{\sum_{i=1}^{n}(y_i-\bar{y})^{2}}{n-2}}$.
  
  n-2 is because we are estimating two parameters ($\beta_0$ and $\beta_1$). This shows that as n increases the estimate of $\sigma$ becomes more precise. As a larger sample provides more information about the variability of the errors making the residual standard error more accurate of an estimate of the true $\sigma$.

2.  Consider a simple linear regression model. State the definition of the $R^2$ statistic by providing also the definitions of RSS and TSS, and explain its connection with sample correlation and the LSE of the slope.

  RSS is the Residual Sum of Squares which measures the amount of variability in $Y$ that isn't explained by the regression: $RSS = \sum_{i=1}^{n}(y_i - \hat{y_i})^{2}$. This equation has $\hat{y}_i = \hat{\beta_0} + \hat{\beta_1}x_i$.
  
  TSS is the Total Sum of Squares which measures the total variability in the observed $Y$'s around their mean: $TSS = \sum_{i=1}^{n} (y_i - \bar{y_i})^{2}$. This measures how much the observed $y_i$'s vary around their sample mean $\bar{y}$.
  
  $R^2$ statistic is $R^2 = 1 - \frac{RSS}{TSS}$ or can be also be written as $R^2 = \frac{TSS-RSS}{TSS}$. This represents the proportion of variability in $Y$ that is explained by the linear regression model.Simple linear regression is equal to the square of the sample correlation between $X$ and $Y$: $R^2=r^2$
  
  This makes $R^2 = [\frac{\sum_{i=1}^{n}(x_i - \bar{x_i})(y_i - \bar{y_i})}{\sqrt{\sum_{i=1}^{n}(x_i - \bar{x_i})^2\sum_{i=1}^{n}(y_i - \bar{y_i})^2}}]^2$. R then describes the strength of the linear relationship but doesn't tell whether the relationship is positive or negative.
  
  LSE of slope: $\hat{\beta_1} = \frac{\sum_{i=1}^{n}(x_i-\bar{x})(y_i-\bar{y})} {\sum_{i=1}^{n}(x_i-\bar{x})^2}$. The numerator of $\hat{\beta_1}$ is the cross-product that appears in the sample correlation $r$. This shows the direct relationship between the slope and the correlation is $\hat{\beta_1} = r\frac{s_Y}{s_X}$. $s_X$ is the standard deviation of $X$ and $s_Y$ is the standard deviation of $Y$.
  
  If $r>0$ then $\hat{\beta_1}>0$. If $r<0$ then $\hat{\beta_1}<0$. If $r=0$ then $\hat{\beta_1}=0$. But $R^2$ only measures the strength of thelinear relationship but not its directions sho it cannot tell us about the direction of $\hat{\beta_1}$.

3.  Consider a multiple linear regression model $$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_p X_p + \varepsilon,$$ where $E(\varepsilon)=0$ and $Var(\varepsilon)=\sigma^2$, $\beta_0$ is the intercept, and $\beta_1,\cdots,\beta_p$ are the regression coefficients. Suppose we want to test that a particular subset of $q$ of the $p$ regression coefficients are zero.

- 

  (a) Provide a test statistic for this by describing in detail each quantity that makes up the statistic.

  $H_0: \beta_{p-q+1} = \beta_{p-q+2} = \cdots = \beta_{p} = 0$
  
  $H_A:$ at least one of the q coefficients isn't zero.
  
  An appropriate test to use for this is the $F$-test, where $F=\frac{(RSS_0-RSS)/q}{RSS/(n-p-1)}$. $RSS$ is the residual sum of squares for the full model that contains all $p$ predictors ($Y = \beta_0 + \beta_1X_1 + \cdots + \beta_pX_p + \varepsilon$). $RSS_0$ is the residual sum of squares for the model that uses all the variables except the last $q$.
  
  Under $H_0$, $F\sim F_{q,n-p-1}$. $F$-statistic has $q$ and $n-p-1$ degrees of freedom. In this test $\frac{RSS_0-RSS}{q}$ measures how much the $RSS$ decreases per coefficient when we add the $q$ predictors., and $\frac{RSS}{n-p-1}$ is the estimatae of the error variance $\sigma^2$. Thus, if the $q$ predictors provide little additional explanatory power, the $F$-statistic should be relatively small.

  (b) Under what conditions will the test statistic work well?
  
  Linear - 

4.  Question 3(a) of Section 3.7 (the section for exercises) of the Textbook. 3(a)

# Applied exercises

1.  Exercise 9 of Section 3.7 of the Textbook.

<!-- -->

9)  This question involves the use of multiple linear regression on the Auto data set.

<!-- -->

(a) Produce a scatterplot matrix which includes all of the variables in the data set.

```{r}
library(ISLR)
pairs(Auto)
```

(b) Compute the matrix of correlations between the variables using the function cor(). You will need to exclude the name variable, which is qualitative.

```{r}
auto_corr_matrix <- cor(Auto[c(1:8)])
auto_corr_matrix
```

(c) Use the lm() function to perform a multiple linear regression with mpg as the response and all other variables except name as the predictors. Use the summary() function to print the results. Comment on the output.

```{r}
lm.fit1 <- lm(mpg ~ . - name, data = Auto)
summary(lm.fit1)
```

i.  Is there a relationship between the predictors and the response?

Yes, there's a relationship between the predictors and the response. Testing the null hypothesis of whether all the regression coefficients are zero shows this as the F-statistic is far from 1, with a very small p-value, showing evidence against the null hypothesis.

ii. Which predictors appear to have a statistically significant relationship to the response?

By looking at the p-values associated with each predictor's t-statistic the displacement, weight, year, and origin have a statistically significant relationship, while cylinders, horsepower, and acceleration don't.

iii. What does the coefficient for the year variable suggest?

The regression coefficient for the year suggests that for every one year mpg increases by 0.750773. This shows how car are becoming more fuel efficient every year by 0.750773mpg/year

<!-- -->

(d) Use the plot() function to produce diagnostic plots of the linear regression fit. Comment on any problems you see with the fit.Do the residual plots suggest any unusually large outliers? Does the leverage plot identify any observations with unusually high leverage?

```{r}
par(mfrow = c(2, 2))
plot(lm.fit1)
```

(e) Use the $*$ and $:$ symbols to fit linear regression models with interaction effects. Do any interactions appear to be statistically significant?

```{r}
lm.fit2 <- lm(mpg ~ cylinders * displacement, data = Auto) 
summary(lm.fit2)
lm.fit3 <- lm(mpg ~ displacement : horsepower , data = Auto) 
summary(lm.fit3)
```

VS.

```{r}
lm.fit4 <- lm(mpg ~ displacement * horsepower , data = Auto) 
summary(lm.fit4)
```

(f) Try a few different transformations of the variables, such as $\log(X)$, $\sqrt X$, $X^2$. Comment on your findings.

```{r}

```

2.  Exercise 10 of Section 3.7 of the Textbook.

<!-- -->

10) This question should be answered using the Carseats data set.

<!-- -->

(a) Fit a multiple regression model to predict Sales using Price, Urban, and US.

```{r}
library(ISLR)
lm.fit1 = lm(Sales ~ Price + Urban + US, data= Carseats)
summary(lm.fit1)
```

(b) Provide an interpretation of each coefficient in the model. Be careful—some of the variables in the model are qualitative!

(c) Write out the model in equation form, being careful to handle the qualitative variables properly.

(d)  For which of the predictors can you reject the null hypothesis $H_0: \beta_j = 0$?

(e)  On the basis of your response to the previous question, fit a smaller model that only uses the predictors for which there is evidence of association with the outcome.

```{r}
lm.fit2 = lm(Sales ~ Price + US, data = Carseats)
summary(lm.fit2)
```

(f) How well do the models in (a) and (e) fit the data?

(g) Using the model from (e), obtain 95% confidence intervals for the coefficient(s).

```{r}
confint(lm.fit2)
```

(h) Is there evidence of outliers or high leverage observations in the model from (e)?

```{r}
par(mfrow=c(2,2))
plot(lm.fit2)
```

3.  Exercise 14 of Section 3.7 of the Textbook.

<!-- -->

14) This problem focuses on the collinearity problem.

<!-- -->

(a) Perform the following commands in R:

```{r}
set.seed(1)
x1 = runif (100)
x2 = 0.5*x1 + rnorm (100)/10
y = 2 + 2*x1 + 0.3*x2 + rnorm (100)
```
The last line corresponds to creating a linear model in which y is a function of x1 and x2. Write out the form of the linear model. What are the regression coefficients?


(b) What is the correlation between x1 and x2? Create a scatterplot displaying the relationship between the variables.

```{r}

```

(c) Using this data, fit a least squares regression to predict y using 1 and x2. Describe the results obtained. What are $\hat{\beta_0}$, $\hat{\beta_1}$, and $\hat{\beta_2}$? How do these relate to the true $\beta_0$, $\beta_1$, and $\beta_2$? Can you reject the null hypothesis $H_0: \beta_1 = 0$? How about the null hypothesis $H_0: \beta_2 = 0$?

(d) Now fit a least squares regression to predict y using only x1. Comment on your results. Can you reject the null hypothesis $H_0: \beta_1 = 0$?

```{r}

```

(e) Now fit a least squares regression to predict y using only x2.Comment on your results. Can you reject the null hypothesis $H_0: \beta_1 = 0$?

```{r}

```

(f) Do the results obtained in (c)–(e) contradict each other? Explain your answer.

(g) Now suppose we obtain one additional observation, which was unfortunately mismeasured.

```{r}
  x1=c(x1, 0.1)
  x2=c(x2, 0.8)
  y=c(y,6)
```
Re-fit the linear models from (c) to (e) using this new data. What effect does this new observation have on the each of the models? In each model, is this observation an outlier? A high-leverage point? Both? Explain your answers.

